---
title: "Exercise 3: Table 1"
author: "YOUR NAME HERE"
date: "`r format(Sys.time(), '%m-%d-%Y', tz = 'America/Los_Angeles')`"
params:
  week: "week_9"
output:
  word_document: default
---

\newpage

# Rmd setup

```{r setup, include=FALSE}
# Libraries
library(tidyverse)
library(flextable)
library(openxlsx)
library(here)
library(gt)
library(gtsummary)

knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Instructions

## Submission 

Please knit your file as a **HTML File** and submit your assignment to **bCourses**.

## Tips 

  - The question in **Exercise 3** is meant to complement `Module: Creating Table 1 in R` in `Week 9`. 
  
  - **Exercise 3** is going to use the `glp1_data.csv` dataset that we cleaned in `Module: Creating Table 1 in R`.

\newpage

# Read in Data

```{r}
# Them, import the dataset
data <- read_csv(here(params$week, "data", "glp1_data.csv"))
```

\newpage

# Question 

In `Module: Creating Table 1 in R` we walked through the separate pieces that go into creating a Table 1. Now it is your turn to put everything together and create your own Table 1! Hint - please use the pieces from `09_table-1.Rmd` to put your table together :)

Please include the following pieces in your Table 1

  - Setting proper labelling (`label`) of the characteristics 
  - Setting the (`type`), statistics (`statistic`), and digits (`digits`) as you wish
  - Add an overall column (`add_overall`)
  - Add a p-value column (`add_p`)
  - Bold significant p-values (`bold_p`)
  - Use a `theme_gtsummary_journal` and/or `theme_gtsummary_compact` theme

```{r}
# Table 1 code here
```


