---
title: 'Lab Module - Dashboards with RMarkdown'
subtitle: 'PHW251B - Data Visualization for Public Health'
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
params:
  week: week_11
---


```{r, include=FALSE}
# Before getting started please install these packages
install.packages("ggplot2")
install.packages("plotly")
```


```{r, include=FALSE}
# Then check their versions "ggplot2" & "plotly"
packageVersion("ggplot2") # should be ‘3.5.1’
packageVersion("plotly") # should be ‘4.10.4’
```


```{r setup, include=FALSE}
# Load required libraries
library(here)   
library(readr)
library(dplyr)
library(flexdashboard) # dashboard versions of R Markdown reports
```


```{r read data, include=FALSE}
# Them, import the rds dataset using `read_rds`
data <- read_rds(here(params$week, "data", "ebola_simulated.rds"))
```


# Overview (First Page)

## Column 1 {data-width=550}

Welcome back to PHW251B! Today we're covering dashboards with RMarkdown and `flexdashboard`. The layout of this module is a bit different than what we're normally used to, because this is a ready-to-knit dashboard! I recommend working through each step of the code to see what it does and then under Knit select "Knit to flex_dashboard".

This example is based on the assigned chapter from the Epidemiologist R Handbook, which can be found here: <https://epirhandbook.com/en/dashboards-with-r-markdown.html>


Note the different arguments in lines 1-8 at the top of this RMD, which specify that the output when you knit this document will be a flex_dashboard organized in columns. In a flexdashboard RMD, the hashes (#) we have previously used to create HTML headers now dictate different parts of the dashboard:

  # Creates a new page
  
  ## Creates a column or a row (columns/rows are determined by line 6!)
      These can take an additional argument for width, as seen in this document.
      
  ### Creates a subsection within that column or row


### Hospital Location

The line above will create a subsection within the first column. Now we just need to decide what to put in it. In keeping with our recent weeks on GIS, let's put a map!

```{r hospital location}
# Load 'sf' & 'tmap' libraries
library(sf)
library(tmap)          

# Group dataset by hospital
data_hospital <- data %>%
  group_by(hospital) %>%
  # Number of observations in each group summed = count
  summarise(Count = n(),
            # Median age of patients
            "Median Age" = round(mean(age, na.rm=T), 1),
            lat = mean(lat, na.rm = T),
            lon = mean(lon, na.rm = T)) %>%
  # Convert to spatial object
  sf::st_as_sf(coords = c("lon", "lat"), crs = 4326)
  
# Here I'm using a different library than you may have seen for mapping in R: tmaps!
# tmaps is a quick and easy-to-use interactive mapping library that interfaces with sf objects

# Set tmap mode to interactive
tmap_mode("view")

# Call a basemap, note that it's the first layer b/c it's on the bottom of the map!
tm_basemap("https://{s}.basemaps.cartocdn.com/rastertiles/voyager_nolabels/{z}/{x}/{y}{r}.png", group = NULL) +
  
# Print result
tm_shape(data_hospital) +
  tm_dots(col = "hospital", size = "Count")
```


## Column 2 {data-width=450}

### Epidemic curve by age

This line above creates the first subsection within the second column. Let's put a chart here:

```{r epi curve}
# Load 'ggplot2' & 'plotly' libraries
library(ggplot2)
library(plotly)   

# Create epidemic curve by age
ggplotly(ggplot(data %>% dplyr::filter(is.na(age_cat) == F), 
                aes(x = date_onset, 
                    fill = age_cat, 
                    color = age_cat, 
                    text = paste("Age Category:", age_cat))) + 
  geom_histogram(binwidth = 15, alpha = 0.75) +
  scale_fill_manual(values = c("#fde725",
                                "#b5de2b",
                                "#6ece58",
                                "#35b779",
                                "#1f9e89",
                                "#26828e",
                                "#31688e",
                                "#3e4989",
                                "#482878",
                                "#440154")) + 
  scale_color_manual(values = c("#fde725",
                                "#b5de2b",
                                "#6ece58",
                                "#35b779",
                                "#1f9e89",
                                "#26828e",
                                "#31688e",
                                "#3e4989",
                                "#482878",
                                "#440154")) + 
  theme(panel.background = element_rect(fill = NA),
        panel.grid = element_line(linetype = 3, color = "grey90")) +
    labs(x = "\nDate of onset",
         y = "Count\n",
         fill = "Age Category\n",
         color = "Age Category\n"), tooltip = "text")
```


### Cases by Hospital 

Here is another subsection for the second column! We can add whatever content we desire for our dashboard: in this case, a table summarizing the number of cases by hospital.

```{r cases by hospital}
# Pipe dataframe into group-by, by hospital
data %>%
  group_by(hospital) %>%
  # Count # of observations in each hospital group
  summarise(Cases = n()) %>%
  # Rename column to be uppercase
  rename("Hospital" = "hospital") %>%
  # Sort such that cases are descending
  arrange(desc(Cases)) %>%
  # Pipe to kable -- prettier table!
  knitr::kable()

```


# Data Explorer (Second Page)

The line above adds a second page under a new tab! Let's visualize the dataset as an interactive table for viewers to explore:

```{css, echo=FALSE}
.datatables.html-widget.html-widget-static-bound { 
  height: auto !important;
  width: 90vw !important;
}
.dataTables_scrollBody {
  height: unset !important;
}
```

```{r datatable, fig.length = 10}
# Load 'DT' library
library(DT) 

datatable(data %>% select(hospital, date_hospitalisation, outcome, gender, age, bmi, fever), 
          #add CSS class for theming
          class = 'cell-border stripe',
          #custom column name
          colnames = c("Hospital", "Date of Hospitalization", "Outcome", "Gender", "Age", "BMI", "Fever"),
          #remove numbered index
          rownames = F,
          #add buttons to export the data within the html (cool!)
          #note that the `dom` argument can accept many, check out documentation for more
          #by calling `Btp`, we keep buttons (B), the table itself (t), and pagination control (p) 
          extensions = 'Buttons', options = list(dom = 'Btpl',
                                                 buttons = c('copy', 
                                                             'csv', 
                                                             'excel', 
                                                             'pdf', 
                                                             'print'),
                                                 #how many entries per page
                                                 pageLength = 10,
                                                 #options for "Show how many entries"
                                                 lengthMenu = c(10, 20, 30)),
          height = "fit-content") %>%
  #bold the `city_ascii` column
  formatStyle("hospital", fontWeight = "bold") %>%
  #round the `aqi` column to one decimal place
  formatRound("bmi", 1)

```

And that's the end of this module -- in essence, it's just a new way to organize and use an RMD to output a different product. And there are many more customizations and additional arguments that can further enhance your dashboard product, so be sure to take a look at the assigned chapter to learn about the possibilities! The final step is to click "Knit to flex_dashboard". This will give you an HTML document that you can share with others.
