---
title: "Module: Table 1 - Baseline Tables"
author: "Ilya Moskalenko"
date: "`r Sys.Date()`"
params:
  week: "week_9"
output:
  html_document:
    df_print: paged
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Libraries
library(tidyverse)
library(flextable)
library(openxlsx)
library(here)
library(gt)
library(gtsummary)
```

# Introduction

- We will build a baseline Table 1 using the `gtsummary` package. This is a well documented and industry standard package for creating Baseline tables in R. You can see the documentation here [link](https://www.danieldsjoberg.com/gtsummary//index.html).

# GLP 1 Dataset from Week 3

- Simulated dataset for an observational electronic health record study evaluating whether `GLP-1 agonists are protective against developing heart disease`
- This dataset mimics the type of data that you would encounter in a healthcare setting.  
The `inclusion criteria` for this cohort was having a diagnosis of 2 diabetes, actively using `insulin`, and being aged 65+ at baseline.

# GLP 1 Agonist Background

- `GLP-1 agonists` are a class of medications that are clinically used to manage blood sugar (glucose) levels in people with Type 2 diabetes and can also help treat obesity.
- The most commonly known GLP-1 agonist is `Ozempic`. 

# Data Dictionary

Here is the `data dictionary` containing the descriptions of each column.

| `Column`            |	`Type`    | `Description`                                                                                  |
|---------------------|-----------|------------------------------------------------------------------------------------------------|
| `patient_id`        |	Character	| Unique identifier for each patient                                                             |
| `glp1`              |	Factor	  | [Exposure] GLP-1 receptor agonists (Ozempic, Wegovy, etc) [Yes, No]                            |
| `heart_disease`     |	Factor	  | [Outcome] Heart disease [Yes, No]                                                              |
| `age`               |	Numeric	  | Age at study entry [between 65–85 years]                                                       |
| `age_category`      |	Factor	  | Age categorized by 5 year intervals [65, 70, 75, 80, 85]                                       |
| `race`              |	Factor	  | Race category [White, Hispanic, Asian, Black, Multiracial, Native American, Pacific Islander]  |
| `gender`            |	Factor	  | Gender identity [Male, Female, Non-binary]                                                     |
| `medicare_plan`	    | Factor	  | Medicare insurance plan type [Original, Advantage, Supplement]                                 |
| `smoking_status`    |	Factor	  | Smoking status at baseline [Never smoker, Former smoker, Current smoker]                       |
| `diabetes`          |	Factor	  | Has type 2 diabetes [Yes - because everyone in this cohort has diabetes]                       |
| `hypertension`      |	Factor	  | Has hypertension [Yes, No]                                                                     |
| `hyperlipidemia`    |	Factor	  | Has hyperlipidemia [Yes, No]                                                                   |
| `hospitalization`   | Factor	  | Was hospitalized at least once over past 2 years [Yes, No]                                     |
| `bmi`	              | Numeric	  | BMI (Body Mass Index)                                                                          |
| `bmi_category`	    | Factor	  | Categorized BMI [Normal weight, Overweight, Obese]                                             |
| `a1c`               |	Numeric	  | Hemoglobin A1C                                                                                 |
| `sweet_beverages`	  | Factor	  | Average number of sweetened beverages per day                                                  |
| `charlson_score`    | Numeric	  | Charlson comorbidity index score                                                               |
| `insulin`           | Numeric	  | Using insulin [Yes, No]                                                                        |
| `metformin`         | Numeric	  | Using metformin [Yes, No]                                                                      |


# Read in Data

```{r}
# Them, import the dataset
data <- read_csv(here(params$week, "data", "glp1_data.csv"))
```

# Part 1 - `tbl_summary` body

## Stratify exposure/treatment group with `tbl_summary`

A few best practices

  - Since this is aggregate level data do not include individual level data such as `patient_id`
  - Do not include your outcome of interest in a baseline table (Table 1) such as `heart_disease`
  - Do not do not include columns with no variability (1 level columns) like `insulin` and `diabetes` since this is a cohort consisting of people with diabetes receiving insulin 

```{r}
data %>%
  select(-c(patient_id, heart_disease, insulin, diabetes)) %>%
  tbl_summary(by = glp1)
```

## Setting labels (`label`)

Here we are going to customize the labels for the columns names that we are presenting in the table

```{r}
data %>%
  # Use tidyverse workflow to exclude IDs or non-baseline fields
  select(-c(patient_id, heart_disease, insulin, diabetes)) %>%
  tbl_summary(
    # Stratify (comparison groups) — optional
    by = glp1,
    # Customize printed variable labels
    label = list(
      age ~ "Age",
      age_category ~ "Age Category",
      gender ~ "Gender",
      race ~ "Race",
      medicare_plan ~ "Medicare Plan",
      smoking_status ~ "Smoking Status",
      hypertension ~ "Hypertension",
      hyperlipidemia ~ "Hyperlipidemia",
      hospitalization ~ "Hospitalization",
      bmi ~ "BMI",
      bmi_category ~ "BMI Category",
      a1c ~ "A1C",
      sweet_beverages ~ "Sweet Beverages",
      charlson_score ~ "Charlson Score",
      metformin ~ "Metformin"))
```

## Specify gtsummary column types (`type`), statistics (`statistic`), and digits (`digits`)

### `type` - type of variables

`categorical` - variables typically just summarize the frequency (n) and proportion (p) of each level (or category) within a variable.

`continuous` - This option summarizes each continuous variable once, providing a single row of summary statistics (e.g., mean ± SD OR median [IQR])

`continuous2` - This option expands the summary to 2 rows per variable, showing different descriptive statistics (e.g., mean ± SD AND median [IQR])

### `statistic` - the summary statisic to display (linked to `type`)

### `digits` - the number of digits to round summary statistics to (linked to `type` & `statistic`)

### Putting it all together

The `type`, `statistic`, and `digits` parameters go hand in hand with each other that is why we are presenting them together.

`type` = `categorical` -> `statistic` = "{n} ({p})" -> `digits` = **two numbers** c(digits_for_n, digits_for_p)

`type` = `continuous` -> `statistic` = "{n} ({p})" -> `digits` = **two numbers** c(digits_for_mean, digits_for_sd)

`type` = `continuous2` - `statistic` = "{median} [{p25}, {p75}]" -> `digits` = **three numbers** -> c(dig_median, dig_p25, dig_p75)

```{r}
data %>%
  # Use tidyverse pipe workflow to select columns before using subsequent function
  select(-c(patient_id, heart_disease, insulin, diabetes)) %>%
  tbl_summary(
    # Stratify
    by = glp1, 
    # Variable types
    type = list(
      # Specify that all categorical columns should be displayed as categorical values
      all_categorical() ~ "categorical",
      # Specify that all continuous columns should be displayed as a continuous2 two-line summary
      all_continuous()  ~ "continuous2",  
      # Override a specific continuous column (eg - age) to be displayed as a continuous one-line summary
      age ~ "continuous"),  
    # Summary statistic outputs
    statistic = list(
      all_categorical() ~ "{n} ({p})",
      # For continuous2 variables, provide two templates (row 1, row 2)
      all_continuous2() ~ c("{mean} ({sd})", "{median} [{p25}, {p75}]"),
      # For 1 row continuous variables, provide 1 template (like age)
      age ~ "{mean} ({sd})"),
    # Number of digits to round to
    digits = list(
      # Categorical - n and p (as a proportion formatted to percent automatically)
      all_categorical() ~ c(0, 1),
    
      # For all continuous2 rows (mean/sd and median/IQR
        # Row 1 - mean (sd)  -> two numbers
        # Row 2 - median [p25, p75] -> three numbers
      all_continuous2() ~ c(1, 1, 0, 1, 1), # 5 numbers is max
    
      # For age (single-row continuous with mean (sd)) -> two numbers
      age ~ c(1, 1)))
```

## Handling missingness (`missing`=)

The `missing` parameter controls whether to show an extra “Unknown”/missing row for each variable.
  - "no" -> do not show missing counts as a row.
  - "ifany" -> show a missing row only if there are NAs.
  - "always" -> always show a missing row.

```{r}
data %>%
  # Use tidverve pipe workflow to select columns before using subsequent function
  select(-c(patient_id, heart_disease, insulin, diabetes)) %>%
  tbl_summary(
    # Stratify
    by = glp1, 
    missing = "always")
```

# Part 2 - Additional Table Features

## Adding overall column `add_overall`

`add_overall` - adds a column with overall (pooled) summaries across strata

Important parameters

  - `last` = FALSE puts the overall column first (default is last).
  - `col_label` = "Overall" sets the column header.

```{r}
data %>%
  select(-c(patient_id, heart_disease, insulin, diabetes)) %>%
  tbl_summary(by = glp1) %>%
  add_overall(last = FALSE, col_label = "Overall")
```

## Adding footnote

```{r}
data %>%
  # Use tidverve pipe workflow to select columns before using subsequent function
  select(-c(patient_id, heart_disease, insulin, diabetes)) %>%
  tbl_summary(
    # Stratify
    by = glp1) %>%
  # Need to convert to gt table
  as_gt() %>%
  # Footnote
  tab_source_note("Your footnote here")
```

## Adding p-value column `add_p`

### `add_p` - adds hypothesis tests comparing strata for each row.

The `test` parameter allows you to control the statistical test

  - `categorical`: "chisq.test" or "fisher.test" (for small counts).
  - `continuous`/`continuous2`: t-test ("t.test") if normality is plausible; otherwise Wilcoxon ("wilcox.test")
 
The `pvalue_fun` parameter allows you to control the number of digits in p-value

NOTE - “Table 1 fallacy” Avoid over-interpreting p-values in a purely descriptive baseline table. Prefer standardized differences for causal/comparative designs; use p-values only when they serve a clear inferential goal. (I usually do not present this)

```{r}
data %>%
  select(-c(patient_id, heart_disease, insulin, diabetes)) %>%
  tbl_summary(by = glp1) %>%
  add_p(
    test = list(
      all_continuous2() ~ "t.test",
      # all_continuous()  ~ "t.test",  # alternative if using simple continuous
      all_categorical() ~ "chisq.test",
      age ~ "t.test",
      sweet_beverages ~ "t.test"
    ),
    # Example of passing args: use Fisher's test for small tables
    # test = list(smoking_status ~ "fisher.test"),
    pvalue_fun = ~ style_pvalue(.x, digits = 3)
  )
```

## Adding summary statistic labels `add_stat_label` 

`add_stat_label` adds a statement to the row label that literally prints the template used to compute the summary statistic (e.g., “n (%)”, “Mean (SD)”).

```{r}
data %>%
  # Use tidyverse pipe workflow to select columns before using subsequent function
  select(-c(patient_id, heart_disease, insulin, diabetes)) %>%
  tbl_summary(
    # Stratify
    by = glp1, 
    # Variable types
    type = list(
      # Specify that all categorical columns should be displayed as categorical values
      all_categorical() ~ "categorical",
      # Specify that all continuous columns should be displayed as a continuous2 two-line summary
      all_continuous()  ~ "continuous2",  
      # Override a specific continuous column (eg - age) to be displayed as a continuous one-line summary
      age ~ "continuous"),  
    # Summary statistic outputs
    statistic = list(
      all_categorical() ~ "{n} ({p})",
      # For continuous2 variables, provide two templates (row 1, row 2)
      all_continuous2() ~ c("{mean} ({sd})", "{median} [{p25}, {p75}]"),
      # For 1 row continuous variables, provide 1 template (like age)
      age ~ "{mean} ({sd})"),
    # Number of digits to round to
    digits = list(
      # Categorical - n and p (as a proportion formatted to percent automatically)
      all_categorical() ~ c(0, 1),
    
      # For all continuous2 rows (mean/sd and median/IQR
        # Row 1 - mean (sd)  -> two numbers
        # Row 2 - median [p25, p75] -> three numbers
      all_continuous2() ~ c(1, 1, 0, 1, 1), # 5 numbers is max
    
      # For age (single-row continuous with mean (sd)) -> two numbers
      age ~ c(1, 1))) %>%
  add_stat_label()
```

# Part 3 - Table Customization

## Font `opt_table_font`

The `opt_table_font` parameter allows you to control the table font after you convert to a gt table using the `as_gt()` function. You can explore the different Google fonts which you can use with `gt::google_font("Fontname")` or simply using a default fonts such as `"Times New Roman"`. You can then using the `tab_options()` control the font size with the `table.font.size = px(n)` parameter and amount of padding with the `data_row.padding = px(2)` parameter.

```{r}
data %>%
  # Use tidverve pipe workflow to select columns before using subsequent function
  select(-c(patient_id, heart_disease, insulin, diabetes)) %>%
  tbl_summary(
    # Stratify
    by = glp1) %>%
  # Need to convert to gt table
  as_gt() %>%
  opt_table_font(
    font = gt::google_font("Nunito")
    # font = "Times New Roman"
    ) %>%
  tab_options(
    table.font.size = px(14), # change font size
    data_row.padding = px(2) # optional tighter layout
)
```

## Bold significant p-values `bold_p`

```{r}
data %>%
  select(-c(patient_id, heart_disease, insulin, diabetes)) %>%
  tbl_summary(by = glp1) %>%
  add_p(
    test = list(
      all_continuous2() ~ "t.test",
      # all_continuous()  ~ "t.test",  # alternative if using simple continuous
      all_categorical() ~ "chisq.test",
      age ~ "t.test",
      sweet_beverages ~ "t.test"
    ),
    # Example of passing args: use Fisher's test for small tables
    # test = list(smoking_status ~ "fisher.test"),
    pvalue_fun = ~ style_pvalue(.x, digits = 3)
  ) %>%
  bold_p(t = 0.05)
```

## Bold variable labels `bold_labels` (section headers)

```{r}
data %>%
  # Use tidverve pipe workflow to select columns before using subsequent function
  select(-c(patient_id, heart_disease, insulin, diabetes)) %>%
  tbl_summary(
    # Stratify
    by = glp1) %>%
  bold_labels()
```

## Italicize level labels (category names) `italicize_levels`

```{r}
data %>%
  # Use tidverve pipe workflow to select columns before using subsequent function
  select(-c(patient_id, heart_disease, insulin, diabetes)) %>%
  tbl_summary(
    # Stratify
    by = glp1) %>%
  italicize_levels()
```

# Part 4 - Table Themes

| Theme Function                                      | Description                                                              | Common Use                            |
| --------------------------------------------------- | ------------------------------------------------------------------------ | ------------------------------------- |
| `theme_gtsummary_journal("jama")`                   | JAMA-style layout (bold headers, compact decimal formatting).            | Publication prep for JAMA.            |
| `theme_gtsummary_journal("nejm")`                   | NEJM-style with clean grid lines, minimal shading.                       | Clinical manuscripts.                 |
| `theme_gtsummary_journal("lancet")`                 | Follows *The Lancet* conventions (no % signs, p-values < 0.0001).        | Epidemiologic/medical papers.         |
| `theme_gtsummary_compact()`                         | Reduces spacing and padding for dense tables.                            | Reports or slides with limited space. |

## JAMA
```{r}
theme_gtsummary_journal(journal = "jama")

data %>%
  select(-c(patient_id, heart_disease, insulin, diabetes)) %>%
  tbl_summary(by = glp1) %>%
  add_overall() %>%
  add_p() 
```

## JAMA
```{r}
theme_gtsummary_journal(journal = "jama")
theme_gtsummary_compact()

data %>%
  select(-c(patient_id, heart_disease, insulin, diabetes)) %>%
  tbl_summary(by = glp1) %>%
  add_overall() %>%
  add_p() 
```

## NEJM + Compact
```{r}
theme_gtsummary_journal(journal = "nejm")
theme_gtsummary_compact()

data %>%
  select(-c(patient_id, heart_disease, insulin, diabetes)) %>%
  tbl_summary(by = glp1) %>%
  add_overall() %>%
  add_p()
```

## JAMA + Compact
```{r}
theme_gtsummary_journal(journal = "jama")
theme_gtsummary_compact()

data %>%
  select(-c(patient_id, heart_disease, insulin, diabetes)) %>%
  tbl_summary(by = glp1) %>%
  add_overall() %>%
  add_p()
```

## Lancet + Compact
```{r}
theme_gtsummary_journal(journal = "lancet")
theme_gtsummary_compact()

data %>%
  select(-c(patient_id, heart_disease, insulin, diabetes)) %>%
  tbl_summary(by = glp1) %>%
  add_overall() %>%
  add_p()
```

# Part 5 - Exporting Table

## gtsummary table

```{r}
# Create table 1 object
t1 = data %>%
  select(-c(patient_id, heart_disease, insulin, diabetes)) %>%
  tbl_summary(by = glp1) %>%
  add_overall() %>%
  add_p() 

# Convert the gtsummary table to a dataframe or tibble
t1_tib <- as_tibble(t1)

# Look at it
head(t1_tib)

# Save the workbook to a file
write_xlsx(t1_tib, here(params$week, "module", "output", "table_1_demo.xlsx"))
```

## Note about gt

Export gtsummary parts of table and then manually add in features from gt package. gt package tables are difficult to export and coding it takes longer than manually fixing in Microsoft Excel or Google Sheets. 

