---
title: 'Exercise 1 - Bar Charts'
subtitle: 'PHW251B: Data Visualization for Public Health'
author: "YOUR NAME HERE"
date: "`r format(Sys.time(), '%m-%d-%Y', tz = 'America/Los_Angeles')`"
params:
  week: "week_3"
output: pdf_document
header-includes:
  - \usepackage{etoolbox}
  - \pretocmd{\section}{\ifnum\value{section}>0\clearpage\fi}{}{}
---


```{r setup, echo = FALSE, warning=FALSE, message=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, tidy.opts = list(width.cutoff = 85))

# It is good practice to always load your libraries first!
library(tidyverse)
library(ggthemes)
library(scales)
library(RColorBrewer)
```

**Submission** - Please knit your file as a **PDF** and submit your assignment to **Gradescope**.

**Tips** 

  - **Questions 1-4** pertain to content from `Lab Module 0` in `Week 1`.

  - **Questions 5-8** pertain to content from the `Bar Charts Module` in `Week 3`.

  - Please make sure to double check that your code does not run off the page in your knitted **PDF**. To fix this issue, go back to your code, and add new lines in your code by hitting enter so that part of the code chunk is on another line. 

# Question 1 - Perform Copy and Rename

Create a copy of this file to use for your exercise using `Save As` and name the copy as follows `03_exercise-1.Rmd` -> `03_exercise-1-firstname_lastname.Rmd`. 

---

# Question 2 - Why Copy and Rename?

**Select all that TRUE** regarding why we require you to save a copy and rename `Rmd` files with the following suffix `-firstname_lastname` in DataHub? (Eg - [X] this answer is TRUE / [ ] this answer is FALSE).

[ ] You should do this for any Rmd in `Datahub` in this course that you wish to edit and not just exercises and assignments.
[ ] For the instructors to more easily identify you when grading your work.
[ ] It reduces the compute that you use in `DataHub`. 
[ ] To make you do more work.
[ ] To ensure that your version of the code is fully preserved in its static state when the instructors push updates or release new weekly materials in `DataHub`

---

## Question 3 - YAML `week` Parameter

Create a `week` object in your environment by pulling in the week parameter from the `YAML` (Hint - use `$`).

```{r}
# Type your answer here

```

---

## Question 4 - Read in Data

Load in the `here` package and use it to read in `glp1_data.csv` from the `PHW251B/week_3/data`. Name the dataframe `data`.

```{r}
# Type your answer here

```

---

## Question 5 - Data-Ink Ratio 

Describe what the Data-Ink Ratio is in 1 or 2 sentences.

_Type your answer here_

## Question 6 - Create Bar Chart in 1 Call

In `Module: Bar Charts` the `03_barcharts.Rmd` file demonstrates how to customize individual `ggplot2` layers when creating a bar chart. 

Now, using the `data` that we read in for Question 4, put the individual layers (i.e. Captions/Labels, Color, and Legend) together to produce a professional bar chart with a quality Data-Ink ratio using a single call. (Hints - you are going to use a lot of "+" to add layers & use a `theme_*`). Feel free to use any **categorical/factor** variables in the dataset that you want to explore!

```{r}
# Type your answer here

```

---

## Question 7 - Create Stacked Bar Chart

Produce a **stacked bar chart** that shows `medicare plan` on the x-axis cases and the color the stacks by `race` using a diverging palette. Use proper labelling titles, colors, and keep the data-ink ratio in mind!

```{r}
# Type your answer here

```

---

## Question 8 - Create Dodged Bar Chart

You have already created a stacked bar chart for tobacco consumption. Similarly, now that we have gone through grouped bar charts, produce a professional grouped barchart that shows the the number of cases of heart disease cases by each age group's sweetened beverage consumption.

```{r}
# Type your answer here

```
