---
title: "Exercise 1 - Bar Charts"
subtitle: 'PHW251B: Data Visualization for Public Health'
author: "YOUR NAME HERE"
date: "`r format(Sys.time(), '%m-%d-%Y', tz = 'America/Los_Angeles')`"
output:
  word_document: default
params:
  week: week_3
---


```{r setup, echo = FALSE, warning=FALSE, message=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, tidy.opts = list(width.cutoff = 85))

# It is good practice to always load your libraries first!
library(tidyverse)
library(ggthemes)
library(scales)
library(RColorBrewer)
```

# Instructions

## Submission 

Please knit your file as a **Word Document** and submit your assignment to **Gradescope**.

## Tips 

  - **Questions 1-4** pertain to content from `Lab Module 0` in `Week 1`.

  - **Questions 5-8** pertain to content from the `Bar Charts Module` in `Week 3`.

  - Please make sure to double check that your code does not run off the page in your knitted **Word Document**. To fix this issue, go back to your code, and add new lines in your code by hitting enter so that part of the code chunk is on another line. 
  
\newpage  
  
## Data

You are going to be using the same `glp1_data.csv` file as in the `Bar Chart Module`. Here is the data dictionary for that dataset. 

| `Column`            |	`Type`    | `Description`                                                                                  |
|---------------------|-----------|------------------------------------------------------------------------------------------------|
| `patient_id`        |	Character	| Unique identifier for each patient                                                             |
| `glp1`              |	Factor	  | [Exposure] GLP-1 receptor agonists (Ozempic, Wegovy, etc) [Yes, No]                            |
| `heart_disease`     |	Factor	  | [Outcome] Heart disease [Yes, No]                                                              |
| `age`               |	Numeric	  | Age at study entry [between 65–85 years]                                                       |
| `age_category`      |	Factor	  | Age categorized by 5 year intervals [65, 70, 75, 80, 85]                                       |
| `race`              |	Factor	  | Race category [White, Hispanic, Asian, Black, Multiracial, Native American, Pacific Islander]  |
| `gender`            |	Factor	  | Gender identity [Male, Female, Non-binary]                                                     |
| `medicare_plan`	    | Factor	  | Medicare insurance plan type [Original, Advantage, Supplement]                                 |
| `smoking_status`    |	Factor	  | Smoking status at baseline [Never smoker, Former smoker, Current smoker]                       |
| `diabetes`          |	Factor	  | Has type 2 diabetes [Yes - because everyone in this cohort has diabetes]                       |
| `hypertension`      |	Factor	  | Has hypertension [Yes, No]                                                                     |
| `hyperlipidemia`    |	Factor	  | Has hyperlipidemia [Yes, No]                                                                   |
| `hospitalization`   | Factor	  | Was hospitalized at least once over past 2 years [Yes, No]                                     |
| `bmi`	              | Numeric	  | BMI (Body Mass Index)                                                                          |
| `bmi_category`	    | Factor	  | Categorized BMI [Normal weight, Overweight, Obese]                                             |
| `a1c`               |	Numeric	  | Hemoglobin A1C                                                                                 |
| `sweet_beverages`	  | Factor	  | Average number of sweetened beverages per day                                                  |
| `charlson_score`    | Numeric	  | Charlson comorbidity index score                                                               |
| `insulin`           | Numeric	  | Using insulin [Yes, No]                                                                        |
| `metformin`         | Numeric	  | Using metformin [Yes, No]                                                                      |

\newpage

# Question 1 - Perform Copy and Rename

Create a copy of this file to use for your exercise using `Save As` and name the copy as follows `03_exercise-1.Rmd` -> `03_exercise-1-firstname_lastname.Rmd`. 

\newpage

# Question 2 - Why Copy and Rename?

**Select all that TRUE** regarding why we require you to save a copy and rename `Rmd` files with the following suffix `-firstname_lastname` in DataHub? (Eg - [X] this answer is TRUE / [ ] this answer is FALSE).

[ ] You should do this for any Rmd in `Datahub` in this course that you wish to edit and not just exercises and assignments.
[ ] For the instructors to more easily identify you when grading your work.
[ ] It reduces the compute that you use in `DataHub`. 
[ ] To make you do more work.
[ ] To ensure that your version of the code is fully preserved in its static state when the instructors push updates or release new weekly materials in `DataHub`

\newpage

# Question 3 - YAML `week` Parameter

Create a `week` object in your environment by pulling in the week parameter from the `YAML` (Hint - use `$`).

```{r}
# Type your answer here

```

\newpage

# Question 4 - Read in Data

Load in the `here` package and use it to read in `glp1_data.csv` from the `PHW251B/week_3/data`. Name the dataframe `data`. 

```{r}
# Type your answer here

```

\newpage

# Question 5 - Data-Ink Ratio 

Describe what the Data-Ink Ratio is in 1 or 2 sentences.

_Type your answer here_

# Question 6 - Create Bar Chart in 1 Call

In `Module: Bar Charts` the `03_barcharts.Rmd` file demonstrates how to customize individual `ggplot2` layers when creating a bar chart. 

Now, using the `data` that we read in for Question 4, put the individual layers (i.e. Captions/Labels, Color, and Legend) together to produce a professional bar chart with a quality Data-Ink ratio using a single call. (Hints - you are going to use a lot of "+" to add layers & use a `theme_*`). Feel free to use any **categorical/factor** variables in the dataset that you want to explore!

```{r}
# Type your answer here

```

\newpage

# Question 7 - Create Stacked Bar Chart

Produce a **stacked bar chart** that shows `medicare plan` on the x-axis cases and the color the stacks by `race` using a diverging palette. Use proper labeling titles, colors, and us a theme to reduce reduce the data ink ratio.

```{r}
# Type your answer here

```

\newpage

# Question 8 - Create Dodged/Grouped Bar Chart

Amend your code from **Question 7** to produce a **dodged/grouped bar chart** that shows `medicare plan` on the x-axis cases and the color the groups by `race` using a diverging palette. Use proper labeling titles, colors, and us a theme to reduce reduce the data ink ratio.

```{r}
# Type your answer here

```
