---
title: "Exercise 4 - Interactive Plots & Tables"
subtitle: 'PHW251B: Data Visualization for Public Health'
author: "YOUR NAME HERE"
date: "`r format(Sys.time(), '%m-%d-%Y', tz = 'America/Los_Angeles')`"
params:
  week: "week_10"
output:
  html_document:
    df_print: paged
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(here)
```

\newpage

# Instructions

## Submission 

Please knit your file as a **HTML Document** (to display interactivity) and submit your assignment to **bCourses**.

## Tips 

  - The questions in **this exercise** are meant to complement `Lab Module: Interactive Plots & Tables` in `Week 10`.
  
  - **This exercise** is going to use the `ca_aqi.rds` dataset that we cleaned in `Module: Scatterplots`

  - Please make sure to double check that your code does not run off the page in your knitted **HTML Document**. To fix this issue, go back to your code, and add new lines in your code by hitting enter so that part of the code chunk is on another line. 

\newpage

# Read in the Data

```{r import data}
# Week parameter
week = params$week

# Read clean rds (R dataset) file with `write_rds`
ca_aqi = read_rds(here(week, "data", "top5_cities.rds"))
```

\newpage

# Run this code

These steps are going to create dataframe `top5_df` for the 5 cities with highest average AQI in California using `ca_aqi`.

```{r}
# Here we generate the names of cities we are interested in
top5cities <- ca_aqi %>%
  # Group by city
  group_by(city_ascii) %>%
  # Population-weighted average of AQI
  summarise(mean_aqi = weighted.mean(aqi, population, na.rm = T)) %>%
  # Sort dataframe by mean aqi, in descending order
  dplyr::arrange(desc(mean_aqi)) %>%
  # Take the top 5
  head(5) %>%
  # Select just the city names
  select(city_ascii) %>%
  # Convert the dataframe into a vector
  as_vector()

# Create dataframe with all unaggregated AQI measurements
top5_df <- ca_aqi %>%
  # Filter for cities within our list
  filter(city_ascii %in% top5cities,
         # Filter out extreme aqi values
         aqi < 1000)
```

# Run Package Updates and Versions

Before building update "ggplot2" & "plotly"

```{r, message=FALSE}
# Update ggplot2 & update plotly
install.packages("ggplot2")
install.packages("plotly")
```

Then check their versions "ggplot2" & "plotly"

```{r}
# Check package versions (Check package versions)
packageVersion("ggplot2") # should be ‘3.5.1’
packageVersion("plotly") # should be ‘4.10.4’
```


\newpage

# Question 1

Use the `top5_df` dataframe we've prepared above (the 5 cities with highest average AQI in California) from `Lab Module: Interactive Plots & Tables` and create either an interactive **violin plot** or an interactive **box plot** to depict the spread of AQI measurements across the five cities using`ggplot` with a `ggplotly` conversion.

Of importance is to take note as to how the tooltip is being generated. Refer to the code from `Lab Module: Interactive Plots & Tables` to customize the tooltip if you deem it necessary.

```{r}

```


\newpage

# Question 2

Build the same plot as you did in *Question 1* using native `plotly` object construction.

```{r}

```
